Optimal updating factor in Wang-Landau and metadynamics simulations

### 2016-02-04 doc/invt.tex
invt.h adding code for the exact schedule
add errsigscan.gp and alphacmp.gp

If we can assume that before entering the asymptotic regime,
the system has entered the equilibrium under a fixed $\alpha_0$,
we can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s in Eq. \eqref{eq:error_res}.

\note{Another possible approximation
  is to interpret $\Gamma_k$ as twice the autocorrelation time,
  and the kernel is exponential
  $$
  \kappa_k = \exp\left( - \frac{2}{\Gamma_k} |t| \right),
  $$
  then a slightly more accurate value is given by\cite{vankampen}
  $$
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac{      ( \lambda_k \, \alpha_0 )^2     }
       { \lambda_k \, \alpha_0 + 2 / \Gamma_k }.
  $$
  But it appears that the main error of Eq. \eqref{eq:y2_eql}
  is an underestimate instead of an overestimate.
  And a possible source of error might be from
  converting a discrete sum to a continuous integral.
}

\subsubsection{\label{twolimits}
Asymptotically optimal schedule}

The asymptotic error given by Eq. \eqref{eq:error_asym1}
can be rewritten as
%
\begin{equation}
  \Err_A
  =
  \int_0^t
    {\mathcal L} \bigl[ q(\tau)\bigr]
    \, d\tau,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where, we have defined the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(\tau) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl( q(\tau) \bigr)
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl( q(\tau) \bigr)
  \; \dot q^2( \tau ).
  \label{eq:error_asym_lagr}
\end{align}
%
To minimize the functional,
we notice that the Lagrangian given by Eq. \eqref{eq:error_asym_lagr}
does not explicitly depend on time, $t$.
%
Thus, the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
is conserved, and in this case,
the Hamiltonian and the Lagrangian coincide.
%
So we may write
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(\tau) \bigr]
  \;
  \dot q^2(\tau)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating integrate the square root of the left-hand side,
%
\begin{equation}
  \int_0^{ q(\tau) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, d\tau
  ,
  \label{eq:q_opt}
\end{equation}
%
we get an implicit equation for the optimal $\alpha(t')$
via $q(\tau) = \int_0^\tau \alpha(t') \, dt'$.

@@

Generally, for an arbitrary updating matrix $\mathbf w$,
direct optimization of Eq. \eqref{eq:error_asym1}
often leads to a complicated expression for $\alpha(t)$.
%
In practice, it is also useful to find
a simple expression for $\alpha(t)$
that is optimal with respect to some free parameter(s).
%
From Eq. \eqref{eq:alpha_opt},
a reasonable candidate would be
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.

@@

In practice, however,
real simulations
may be relative short
and have not entered
the asymptotic regime.
%
Then, the residual error, $\Err_R$, is not negligible.
%
From Eq. \eqref{eq:error_res}, we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { t + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is given by the sum of
Eqs. \eqref{eq:error_res_invt} and \eqref{eq:error_asym_invt}.
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \Err^f_k \,
  \left\{
    1
    +
    \frac{ 1 - \left[ t_0 / (t + t_0) \right]^{2 \, \nu_k - 1} }
         {                 2 \, \nu_k - 1                      }
%    \frac{       1        }
%         { 2 \, \nu_k - 1 }
%    \left[
%      1 -
%      \left(
%        \tfrac{   t_0   }
%              { t + t_0 }
%      \right)^{2 \, \nu_k - 1}
%    \right]
  \right\}
,
\label{eq:error_invt}
\end{align}
%
where
$$
\Err^f_k
\equiv
\left\langle
  y_k^2
\right\rangle_{ \alpha(t) }
=
\frac{  \Gamma_k \, \nu_k   }
     {    2 \, (t + t_0)    }
,
$$
is the equilibrium error of mode $k$
under the constant terminal value
of $\alpha(t) = 1/[\lambda (t + t_0)]$.

@@

\subsection{Optimality of the single-bin scheme}

Randomize the initial error?

Error vs. $\mu_1$.

Error vs. $\sigma$.

@@

\section{Acknowledgments}

We thanks Dr. Y. Mei and Dr. J. Ma for helpful discussions.

>>>>

If we can assume that before entering the asymptotic regime,
the system has entered the equilibrium under a fixed $\alpha_0$,
we can then use Eq. \eqref{eq:y2_eql}
for the $\langle y_k^2(0) \rangle$'s in Eq. \eqref{eq:error_res}.
%
\note{Another possible approximation
  is to interpret $\Gamma_k$ as twice the autocorrelation time,
  and the kernel is exponential
  $$
  \kappa_k = \exp\left( - \frac{2}{\Gamma_k} |t| \right),
  $$
  then a slightly more accurate value is given by\cite{vankampen}
  $$
  \left\langle
    y_k^2
  \right\rangle
  =
  \frac{      ( \lambda_k \, \alpha_0 )^2     }
       { \lambda_k \, \alpha_0 + 2 / \Gamma_k }.
  $$
  But it appears that the main error of Eq. \eqref{eq:y2_eql}
  is an underestimate instead of an overestimate.
  And a possible source of error might be from
  converting a discrete sum to a continuous integral.
}
%
In this case, we have
%
\begin{equation}
  \Err_R
  =
  \sum_{k = 2}^n
      \frac 1 2 \, \alpha_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(t) }
  .
  \label{eq:error_res1}
\end{equation}
%
It follows that the residue error
depends on the schedule $\alpha(\tau)$
only through the integral $q(t)$.
%
Thus, if we variate the schedule under a fixed $q(t)$,
then this part can be ignored,
and the optimization can be done exclusively
on the asymptotic part, $\Err_A$.

\subsubsection{\label{twolimits}
Optimal schedule}

As shown above, with fixed endpoint value of $q(t)$,
the optimization of the error is equivalent to
that of the asymptotic error.
%
We now rewrite Eq. \eqref{eq:error_asym1} as
%
\begin{equation}
  \Err_A
  =
  \int_0^t
    {\mathcal L} \bigl[ q(\tau)\bigr]
    \, d\tau,
  \notag
  %\label{eq:error_asym_action}
\end{equation}
%
where, we have defined the Lagrangian,
%
\begin{align}
  {\mathcal L} \bigl[ q(\tau) \bigr]
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \dot u_k^2\bigl( q(\tau) \bigr)
  \notag
  \\
  &=
  \sum_{ k = 2 }^n
    \Gamma_k \, \lambda_k^2 \, u_k^2 \bigl( q(\tau) \bigr)
  \; \dot q^2( \tau ).
  \label{eq:error_asym_lagr}
\end{align}
%
To minimize the functional,
we notice that the Lagrangian given by Eq. \eqref{eq:error_asym_lagr}
does not explicitly depend on time, $t$.
%
Thus, the Hamiltonian,
%
\begin{equation}
  \mathcal H
  =
  \frac{ \partial \mathcal L }
       { \partial \dot q     }
  \, \partial \dot q
  -
  \mathcal L
  =
  2 \, \mathcal L
  - \mathcal L
  =
  \mathcal L,
  %=
  %\mathrm{const.}
  \notag
  %\label{eq:error_asym_Hamiltonian}
\end{equation}
%
is conserved, and in this case,
the Hamiltonian and the Lagrangian coincide.
%
So we may write
%
\begin{equation}
  \mathcal L
  =
    \textstyle\sum_{ k = 2 }^n
      \Gamma_k \, \lambda_k^2
      \, u_k^2 \bigl[ q(\tau) \bigr]
  \;
  \dot q^2(\tau)
  =
  c^2
  \quad
  (c > 0).
  \label{eq:Lagrangian_const}
\end{equation}
%
Integrating integrate the square root of the left-hand side,
%
\begin{equation}
  \int_0^{ q(\tau) }
    \sqrt{
      \textstyle\sum_{ k = 2 }^n
        \Gamma_k \, \lambda_k^2
        \, u_k^2( q' )
    }
    \;
    d q'
  =
  c \, d\tau
  ,
  \label{eq:q_opt}
\end{equation}
%
we get an implicit equation for the optimal $\alpha(\tau)$
via its integral $q(\tau)$.

@@

Generally, for an arbitrary updating matrix $\mathbf w$,
direct optimization of Eq. \eqref{eq:error_asym1}
often leads to a complicated expression for $\alpha(t)$.
%
In practice, it is also useful to find
a simple expression for $\alpha(t)$
that is optimal with respect to some free parameter(s).
%
From Eq. \eqref{eq:alpha_opt},
a reasonable candidate would be
%
\begin{equation}
\alpha(\tau) = \frac{1}{\lambda \, (\tau + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
for some value of $\lambda$.
%
Below we shall assume this functional form
and find the optimal value
in some cases.
%
We shall show that the optimal $\lambda$
generally depends on not only the spectrum
of eigenvalues and the sampling method,
but also other parameters,
such as the simulation length, $t$.
%
Note, however,
since Eq. \eqref{eq:alpha_invtlambda}
does not fix the endpoint value,
%
\begin{equation}
q(t) = \frac{ 1 } { \lambda }
\log\left(
  \frac{ t + t_0 } { t_0 }
\right),
\label{eq:qt_invtlambda}
\end{equation}
%
we have to include the residual error
during the optimization.

@@

In practice, however,
real simulations
may be relative short
and have not entered
the asymptotic regime.
%
Then, the residual error, $\Err_R$, is not negligible.
%
From Eqs. \eqref{eq:error_res1} and \eqref{eq:qt_invtlambda},
we get
%
\begin{equation}
\Err_R
=
\sum_{k = 2}^n
  \frac{ \Gamma_k \, \nu_k }
       {        2 \, t_0   }
  \left(
      \frac{   t_0   }
           { t + t_0 }
   \right)^{ 2 \, \nu_k },
\label{eq:error_res_invt}
\end{equation}
%
assuming that at $t = 0$
the system has achieved an equilibrium
at a constant $\alpha_0$,
such that the values of
$\left\langle y_k^2(0) \right\rangle$
can be computed from Eq. \eqref{eq:y2_eql}.
%
The total error is given by the sum of
Eqs. \eqref{eq:error_res_invt} and \eqref{eq:error_asym_invt}.
%
\begin{align}
\Err
&=
\Err_R + \Err_A
\notag
\\
&=
\sum_{ k = 2 }^n
  \Err^f_k \,
  \left\{
    1
    +
    \frac{ 1 - \left[ t_0 / (t + t_0) \right]^{2 \, \nu_k - 1} }
         {                 2 \, \nu_k - 1                      }
%    \frac{       1        }
%         { 2 \, \nu_k - 1 }
%    \left[
%      1 -
%      \left(
%        \tfrac{   t_0   }
%              { t + t_0 }
%      \right)^{2 \, \nu_k - 1}
%    \right]
  \right\}
,
\label{eq:error_invt}
\end{align}
%
where
$$
\Err^f_k
\equiv
\left\langle
  y_k^2
\right\rangle_{ \alpha(t) }
=
\frac{  \Gamma_k \, \nu_k   }
     {    2 \, (t + t_0)    }
,
$$
is the equilibrium error of mode $k$
under the constant terminal value
of $\alpha(t) = 1/[\lambda (t + t_0)]$.

@@

\subsection{Accuracy of the inverse-time schedule}

In the above, we have assumed the inverse-time schedule,
Eq. \eqref{eq:alpha_invtlambda}.
%
In Fig. \ref{fig:alpha_cmp},
we compared the inverse-time schedule
with the optimal $\lambda$,
and the optimal schedule from integrating
Eq. \eqref{eq:q_opt}.
%
The difference was larger for the global sampling scheme.
%
\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/alphacmp.pdf}
  }
  \caption{
    \label{fig:alpha_cmp}
    Optimal schedules, $alpha(t)$, from Eq. \eqref{eq:q_opt},
    versus the inverse-time schedule,
    Eq. \eqref{eq:alpha_invtlambda},
    with the optimal parameter, $\lambda$.
    %
    For fair comparison,
    the $q(t) = \int_0^t \alpha(\tau) \, d\tau$
    were set to be the same
    for both schedules.
  }
\end{center}
\end{figure}

In Fig. \ref{fig:err_sigscan},
we compared the error from the inverse-time schedule
with the optimal $\lambda$,
and that from the asymptotically optimal schedule,
for the Gaussian updating scheme.
%
The errors from the two schedules appeared to be reasonable close,
unless for large width case.
%
This suggests that very long metadynamics simulations
may benefit from optimal schedule,
instead of the simple formula, given by
Eq. \eqref{eq:alpha_invtlambda}.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsigscan.pdf}
  }
  \caption{
    \label{fig:err_sigscan}
    Normalized error, $(t + t_0) \, E$,
    versus the width of the Gaussian updating scheme,
    $\sigma$,
    according to the schedule,
    Eq. \eqref{eq:alpha_invtlambda},
    with the optimal $\lambda$,
    and or the asymptotically optimal schedule.
  }
\end{center}
\end{figure}

%\subsection{Optimality of the single-bin scheme}

The above comparison also shows the asymptotic superiority
of the single-bin scheme.
It is clear that for short times, 
the single-bin (WL) may not be optimal,
but it ultimately wins out as simulation lengthens.

%Randomize the initial error?
%
%Error vs. $\mu_1$.
%
%Error vs. $\sigma$.

@@

%\section{Acknowledgments}
%
%We thanks Dr. Y. Mei and Dr. J. Ma for helpful discussions.
